---
title: "STAT1378 Part I Report"
author: "ADAB GILL"
output:
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
    pandoc_args: ["--citeproc"]
bibliography: "references.bib"
link-citations: true
---

# Abstract

This report examines the relationships among height, weight, gender, and physical activity levels within a sample dataset of men and women aged 26–45.
The analysis utilizes linear regression, t-tests, and chi-square tests to answer questions about linear relationships, mean differences, and associations [@Xie2015].

# Introduction

The study investigates three main questions: 1.
Is there a linear relationship between height and weight?
2.
Does mean height differ between males and females?
3.
Is there an association between gender and physical activity level?

Statistical methods like linear regression, t-tests, and chi-squared tests are employed, aiming to answer these questions rigorously.

# Methods

## Data Description

The dataset includes 1000 observations, each with the following variables: - **ID**: Unique identifier for each participant.
- **Gender**: Male or Female.
- **Height**: Participant height in centimeters.
- **Weight**: Participant weight in kilograms.
- **Physical Activity**: Level of physical activity (None, Moderate, Intense).

## Statistical Tests

-   **Linear Regression**: To investigate the relationship between height and weight.
-   **t-Test**: To compare mean heights of males and females.
-   **Chi-squared Test**: To examine the association between gender and physical activity.

# Analysis and R Code

## Question 1: Linear Relationship between Height and Weight

```{r, tidy = TRUE, echo = FALSE, include = FALSE}
# Run the following libraries
install.packages(c("rmarkdown", "knitr", "tidyverse", 
                   "kableExtra", "bibtex", "tinytex", "dplyr", "readr", 
                   "ggplot2", "knitr", "gridExtra"))
library(rmarkdown)
library(knitr)
library(tidyverse)
library(kableExtra)
library(bibtex)
library(tinytex)
library(dplyr)
library(readr)
library(ggplot2)
library(gridExtra)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
options(tinytex.verbose = TRUE)
```

```{r, linear-model, message=FALSE, fig.align="center", fig.width=6, fig.height=4}
# Load data
data <- read.csv("project.csv")

# Linear regression model
height_weight_model <- lm(weight ~ height, data = data)
summary(height_weight_model)
model_summary <- summary(height_weight_model)

# Extract coefficients table and convert to data frame
coefficients_df <- as.data.frame(model_summary$coefficients)
colnames(coefficients_df) <- c("Estimate", "Std_Error", "t_value", "p_value")

# Display the summary table using kable without complex LaTeX commands
kable(coefficients_df, format = "latex", booktabs = TRUE, 
      caption = "Linear Regression Summary for Height and Weight") %>%
  kable_styling(position = "center")

# Plot with regression line
ggplot(data, aes(x = height, y = weight)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Relationship between Height and Weight",
       x = "Height (cm)", y = "Weight (kg)") +
  theme_minimal(base_size = 10)  # Ensures readability in PDF
```

## Results for Linear Regression

Based on the linear regression analysis, a significant relationship was found between height and weight, with a p-value of `r summary(height_weight_model)$coefficients[2,4]`.
The slope of the line suggests that with each centimeter increase in height, weight changes by approximately `r round(coef(height_weight_model)[2], 2)` kg [@Moore2016].

## Question 2: Difference in Mean Height between Males and Females

```{r, tidy = TRUE}
# Perform t-test for mean height comparison
t_test_result <- t.test(height ~ gender, data = data)

# Create a data frame with t-test results
t_test_df <- data.frame(
  Statistic = t_test_result$statistic,
  Parameter = t_test_result$parameter,
  p_value = t_test_result$p.value,
  Confidence_Interval = paste(round(t_test_result$conf.int[1], 2), 
                              "-", round(t_test_result$conf.int[2], 2)),
  Mean_Female = t_test_result$estimate[1],
  Mean_Male = t_test_result$estimate[2]
)

# Display table with kable
kable(t_test_df, caption = "t-Test Summary for Mean Height Comparison") %>%
  kable_styling("striped", full_width = FALSE)
```

```{r height-boxplot-density, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}

# Boxplot for mean height by gender
ggplot(data, aes(x = gender, y = height, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Boxplot of Heights by Gender", x = "Gender", y = "Height (cm)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

```


## Results for t-Test

The Welch Two Sample t-test shows a statistically significant difference in mean height between males and females, with a t-value of -39.77 and a p-value less than 2.2e-16.
This result suggests that the mean height of males (178.13 cm) is significantly greater than the mean height of females (167.98 cm).
The 95% confidence interval for the difference in means is between -10.65 and -9.65 cm, indicating that males are, on average, taller than females in this sample.
Since the p-value is much smaller than 0.05, we reject the null hypothesis and conclude that there is a statistically significant difference in height by gender.
The negative t-value and confidence interval (both in the negative range) confirm that males are, on average, taller than females.
The interval provides an estimated range for the difference in means, reinforcing that males are taller on average [@Agresti2018].

The hypotheses for the chi-squared test are:

$$
H_0: \text{Gender and Physical Activity Level are independent}
$$

$$
H_1: \text{Gender and Physical Activity Level are not independent}
$$

## Question 3: Association between Gender and Physical Activity Level

```{r}
# Create a contingency table for Gender and Physical Activity
activity_gender_table <- table(data$gender, data$phys)

# Perform chi-squared test
chi_square_result <- chisq.test(activity_gender_table)

# Create a data frame for chi-squared test results
chi_square_df <- data.frame(
  Statistic = chi_square_result$statistic,
  Degrees_of_Freedom = chi_square_result$parameter,
  p_value = chi_square_result$p.value
)

# Display table with kable
kable(chi_square_df, caption = "Chi-Squared Test Summary for Gender and Physical Activity") %>%
  kable_styling("striped", full_width = FALSE)
```

```{r gender-activity-barplot, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
# Stacked bar plot for physical activity by gender
barplot <- ggplot(data, aes(x = gender, fill = phys)) +
  geom_bar(position = "fill") +  # "fill" makes it proportional within each gender
  labs(title = "Proportion of Physical Activity Levels by Gender",
       x = "Gender", y = "Proportion", fill = "Physical Activity") +
  scale_y_continuous(labels = scales::percent) +  # Show y-axis as percentages
  theme_minimal(base_size = 12)

# Display the bar plot
barplot
```

## Results for Chi-squared Test

The chi-squared test for association between gender and physical activity level shows a chi-squared value of 3.23 with 2 degrees of freedom and a p-value of 0.1993.
Since the p-value is greater than 0.05, we fail to reject the null hypothesis.
This suggests that there is no statistically significant association between gender and physical activity level in this sample.

The p-value of 0.1993 is above the typical significance level of 0.05, indicating that any observed differences in physical activity level by gender are likely due to random chance.
Failing to reject the null hypothesis implies that gender does not have a significant association with physical activity level in this data set.

# Conclusion

1.  **Height and Weight Relationship**: The linear regression analysis indicates a statistically significant relationship between height and weight.
    This suggests that taller individuals in the sample tend to have higher weights, supporting a positive association between these two variables.

2.  **Gender Differences in Height**: The t-test results show a statistically significant difference in mean height between males and females, with males being taller on average.
    This finding is consistent with general observations of height differences by gender.

3.  **Gender and Physical Activity Level**: The chi-squared test for association between gender and physical activity level does not indicate a statistically significant relationship.
    This implies that, within this sample, physical activity levels do not differ significantly between males and females.

# References
